Variance reduction for MCMC methods via martingale representations

### 2020-02-20 aap/Revision/vr-mcmc-revision-new.tex
1st theorem proof update

which coincide with the left side of \ref{eq:parseval}. Due to Parseval identity $\Psi_{q+1,n}(y) \rightarrow 0, n \rightarrow \infty$ for any $y \in \rset^d$ and $\psi_n(y) < \mathsf{E}[|f(X^y_{q,q+1}) - \mathsf{E} f(X^y_{q,q+1})|^2] \leq \frac{\|f\|^2_{\infty}}{4}$. Hence, by Lebesgue dominated convergence theorem, $\mathsf{E}\psi^2_n(X_q^x) \rightarrow 0$ and since $\mathsf{E}[f(X^y_{q,q+1})]$ is a version of the conditional expectation $\mathsf{E}[f(X^x_{q+1}|X^x_q = y]$, it holds in \(L^2(\mathsf{P})\) that

@@

f(X^x_{q+1}) = \mathsf{E}[f(X^x_{q+1})|X^x_{q+1}] + \sum\limits_{k=1}^{\infty}a_{q+1,q+1,k}(X^x_{q})\phi_{k}(\xi_{q+1})

@@

From  numerical point of view another representation of the coefficients \(a_{p,l,k}\)  turns out to be more useful.
\begin{prop}
The coefficients \(a_{p,l,k}\) in \eqref{eq:coeff_mart}  can be alternatively represented as
\begin{eqnarray*}
a_{p,l,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{p,l}\left(\Phi_l(x,\xi)\right)\right]
\end{eqnarray*}
with \(Q_{p,l}(x)=\mathsf{E}\left[f(X^x_{l,p})\right],\) \(p\geq l.\)
%where   a r.v. \(\xi\) independent of \(\mathcal{G}_{l-1}.\)
The functions \((Q_{p,l})_{l=0}^p\)  can be computed by the backward recurrence: $Q_{p,p}(x)=f(x)$ and for $l \in \{0,\ldots,p-1\}$
\begin{eqnarray}
\label{eq:qpl}
Q_{p,l}(x)=\mathsf{E}\left[Q_{p,l+1}(X^x_{l+1})\right].
%=\mathsf{E}[Q_{p,l+1}(\Phi_{l+1}(x,\xi))]
\end{eqnarray}
In the case $\Phi_l=\Phi$ for all $l\geq 1,$ we have
\begin{equation}
\label{eq:definition-bar-a}
\bar a_{r,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{r}\left(\Phi(x,\xi)\right)\right]
\end{equation}
with  \(Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right],\) $r\in \mathbb{N}.$
\end{prop}
\par
Next we show how  the representation \eqref{eq:mart_repr} can be used to  reduce the variance of MCMC algorithms.  Consider the case of time homogeneous transition kernels and introduce a weighted average estimator $\pi_{n}^{N}(f)$ of the form
\begin{equation}\label{eq:29032018a2}
\pi_{n}^{N}(f)=\frac{1}{n}\sum_{p=N+1}^{N+n} f(X^x_{p}),
\end{equation}
where $N\in\mathbb N_0$ is the length of the burn-in period and $n\in\mathbb N$
the number of effective samples.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
\begin{eqnarray}
\nonumber
M_{K,n}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{k=1}^{K}\sum_{l=N+1}^{p} \bar a_{p-l,k}(X_{l-1})\phi_k(\xi_{l})\right]
\\
&=& \sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left(1+\frac{N-l}{n}\right)A_{N+n-l,k}(X_{l-1})\phi_{k}\left(\xi_{l}\right),
\label{eq:29032018a5}
\end{eqnarray}
where
\begin{equation}
\label{eq:definition-A-s,k}
A_{s,k}(x)=\frac{1}{s}\sum_{r=1}^{s}\bar a_{r,k}(x).
\end{equation}
Since \(X_{l-1}\) is independent of \(\xi_{l}\) and \tcr{\(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\neq 0\)}, we obtain
\[
\PE[A_{N+n-l,k}(X_{l-1})\phi_{k}\left(\xi_{l}\right)] = \mathsf{E}[A_{N+n-l,k}(X_{l-1})\CPE{\phi_{k}\left(\xi_{l}\right)}{\mathcal{G}_{l-1}]} = 0
\]
and hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
The representation \eqref{eq:coeff_mart} suggests that the variance of  the  variance-reduced estimator
\begin{equation}
\label{eq:29032018a3}
\pi_{K,n}^N(f)=\pi_n^N(f)-M_{K,n}^N(f)
\end{equation}
should be small for \(K\) large enough. Indeed, since $\PE[ \phi_k(\xi_l) \phi_{k'}(\xi_l)]=0$ if $k \ne k'$, we obtain
\begin{equation}
\label{eq:expression-variance}
\PVar[\pi_{K,n}^N(f)]\leq\sum_{k=K+1}^{\infty}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X_{N+l-1})]
\end{equation}
and $\mathsf{Var}[\pi_{K,n}^N(f)]$ is small provided that the coefficients $A_{s,k}$ decay fast enough as $k\to \infty.$
In the next section   we provide a detailed   analysis of  $\mathsf{Var}[\pi_{K,n}^N(f)]$  for the  ULA  algorithm (see Example~\ref{exam:langevin-algorithm}).
\section{Analysis of variance reduced ULA}
\label{sec:ula_analysis}
In this section we perform the convergence analysis of the ULA algorithm. For the sake of clarity and notational simplicity we restrict our attention to the constant time step, that is, we take $\gamma_k =\gamma$ for any $k \in \nset$.

>>>>

which coincide with the left side of \ref{eq:parseval}. Due to Parseval identity $\Psi_{q+1,n}(y) \rightarrow 0, n \rightarrow \infty$ for any $y \in \rset^d$ and $\psi_n(y) < \mathsf{E}[|f(X^y_{q,q+1}) - \mathsf{E} f(X^y_{q,q+1})|^2] \leq \frac{\|f\|^2_{\infty}}{4}$. Hence, by Lebesgue dominated convergence theorem, $\mathsf{E}\psi^2_n(X_q^x) \rightarrow 0$ and since $\mathsf{E}[f(X^{X^x_q}_{q,q+1})|X^x_{q}] = \mathsf{E}[f(X^{x}_{q+1})|X^x_{q}]$, it holds in \(L^2(\mathsf{P})\) that

@@

f(X^x_{q+1}) = \mathsf{E}[f(X^x_{q+1})|X^x_{q}] + \sum\limits_{k=1}^{\infty}a_{q+1,q+1,k}(X^x_{q})\phi_{k}(\xi_{q+1})

@@

From  numerical point of view another representation of the coefficients \(a_{p,l,k}\)  turns out to be more useful.
\begin{prop}
The coefficients \(a_{p,l,k}\) in \eqref{eq:coeff_mart}  can be alternatively represented as
\begin{eqnarray*}
a_{p,l,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{p,l}\left(\Phi_l(x,\xi)\right)\right]
\end{eqnarray*}
with \(Q_{p,l}(x)=\mathsf{E}\left[f(X^x_{l,p})\right],\) \(p\geq l.\)
%where   a r.v. \(\xi\) independent of \(\mathcal{G}_{l-1}.\)
The functions \((Q_{p,l})_{l=0}^p\)  can be computed by the backward recurrence: $Q_{p,p}(x)=f(x)$ and for $l \in \{0,\ldots,p-1\}$
\begin{eqnarray}
\label{eq:qpl}
Q_{p,l}(x)=\mathsf{E}\left[Q_{p,l+1}(X^x_{l+1})\right].
%=\mathsf{E}[Q_{p,l+1}(\Phi_{l+1}(x,\xi))]
\end{eqnarray}
In the case $\Phi_l=\Phi$ for all $l\geq 1,$ we have
\begin{equation}
\label{eq:definition-bar-a}
\bar a_{r,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{r}\left(\Phi(x,\xi)\right)\right]
\end{equation}
with  \(Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right],\) $r\in \mathbb{N}.$
\end{prop}
\par
Next we show how  the representation \eqref{eq:mart_repr} can be used to  reduce the variance of MCMC algorithms.  Consider the case of time homogeneous transition kernels and introduce a weighted average estimator $\pi_{n}^{N}(f)$ of the form
\begin{equation}\label{eq:29032018a2}
\pi_{n}^{N}(f)=\frac{1}{n}\sum_{p=N+1}^{N+n} f(X^x_{p}),
\end{equation}
where $N\in\mathbb N_0$ is the length of the burn-in period and $n\in\mathbb N$
the number of effective samples.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
\begin{eqnarray}
\nonumber
M_{K,n}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{k=1}^{K}\sum_{l=N+1}^{p} \bar a_{p-l,k}(X^x_{l-1})\phi_k(\xi_{l})\right]
\\
&=& \sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left(1+\frac{N-l}{n}\right)A_{N+n-l,k}(X^x_{l-1})\phi_{k}\left(\xi_{l}\right),
\label{eq:29032018a5}
\end{eqnarray}
where
\begin{equation}
\label{eq:definition-A-s,k}
A_{s,k}(x)=\frac{1}{s}\sum_{r=1}^{s}\bar a_{r,k}(x).
\end{equation}
Since \(X^x_{l-1}\) is independent of \(\xi_{l}\) and \tcr{\(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\neq 0\)}, we obtain
\[
\PE[A_{N+n-l,k}(X^x_{l-1})\phi_{k}\left(\xi_{l}\right)] = \mathsf{E}[A_{N+n-l,k}(X^x_{l-1})\CPE{\phi_{k}\left(\xi_{l}\right)}{\mathcal{G}_{l-1}]} = 0
\]
and hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
The representation \eqref{eq:coeff_mart} suggests that the variance of  the  variance-reduced estimator
\begin{equation}
\label{eq:29032018a3}
\pi_{K,n}^N(f)=\pi_n^N(f)-M_{K,n}^N(f)
\end{equation}
should be small for \(K\) large enough. Indeed, since $\PE[ \phi_k(\xi_l) \phi_{k'}(\xi_l)]=0$ if $k \ne k'$, we obtain
\begin{equation}
\label{eq:expression-variance}
\PVar[\pi_{K,n}^N(f)]\leq\sum_{k=K+1}^{\infty}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X^x_{N+l-1})]
\end{equation}
and $\mathsf{Var}[\pi_{K,n}^N(f)]$ is small provided that the coefficients $A_{s,k}$ decay fast enough as $k\to \infty.$
In the next section   we provide a detailed   analysis of  $\mathsf{Var}[\pi_{K,n}^N(f)]$  for the  ULA  algorithm (see Example~\ref{exam:langevin-algorithm}).
\section{Analysis of variance reduced ULA}
\label{sec:ula_analysis}
In this section we perform the convergence analysis of the ULA algorithm. For the sake of clarity and notational simplicity we restrict our attention to the constant time step, that is, we take $\gamma_k =\gamma$ for any $k \in \nset$.
